---
params:
  fname: "default"
  dir_data: "default"
  exp_type: "slamdunk" # slamdunk or smallslam

title: "SLAMtrack - T>C"
author: "RWH"
date: "`r format(Sys.time(), '%d %B %Y')`"

output: 
  html_document:
    theme: readable 
    highlight: pygments 
    anchor_sections: FALSE

---

```{r setup, include=FALSE}

if (params$exp_type == "slamdunk") {is_slamdunk <- TRUE} else {is_slamdunk <- FALSE}
dir_data <<- params$dir_data  ## <<- to make global variable
exp_type <<- params$exp_type

source(".Rprofile")
source("scripts/SLAMtrack_00_setup.R")
source("scripts/SLAMtrack_11_import_quick.R")

knitr::opts_chunk$set(
  echo = FALSE, 
  error = TRUE, 
  message = FALSE
  # fig.path = paste0(dir_data, "output/") - don't know why this not working
  )

library(kableExtra)

```


# T>C conversion rates

On a linear scale:  

```{r TC_vs_CPM}

df_tcounts %>% 
  ggplot(aes(x = readsCPM, y = conversionRate, colour = group)) +
  geom_point(alpha = 0.01) +
  facet_wrap(~tissue) +
  scale_x_log10() +
  xlab("gene abundance\nreads per million") +
  ylab("T>C conversion rate") -> p

p %>% add_theme_SLAMtrack() 

df_tcounts %>% 
  ggplot(aes(x = readsCPM, y = conversionRate, colour = group)) +
  geom_smooth() +
  facet_wrap(~tissue) +
  scale_x_log10() + 
  xlab("gene abundance\nreads per million") +
  ylab("T>C conversion rate") -> p

p %>% add_theme_SLAMtrack() 

```

On sqrt / log scales:  

```{r TC_vs_CPM_log10}

df_tcounts %>% 
  ggplot(aes(x = readsCPM, y = conversionRate, colour = group)) +
  geom_point(alpha = 0.01) +
  facet_wrap(~tissue) +
  scale_x_log10() +
  scale_y_sqrt() +
  xlab("gene abundance\nreads per million") +
  ylab("T>C conversion rate") -> p

p %>% add_theme_SLAMtrack() 

df_tcounts %>% 
  ggplot(aes(x = readsCPM, y = conversionRate, colour = group)) +
  geom_smooth() +
  facet_wrap(~tissue, labeller = labeller(.default = toupper)) +
  scale_x_log10() +
  scale_y_sqrt() +
  xlab("gene abundance\n(reads per million)") +
  ylab("T>C conversion rate\n(measure of RNA labelling)") -> p

p %>% add_theme_SLAMtrack() %>% remove_guide_fill()


df_tcounts %>% 
  mutate(conversionRate = conversionRate + cR_nudge) %>% 
  ggplot(aes(x = conversionRate, y = 1, colour = group)) +
  geom_density_ridges(fill = NA) +
  facet_wrap(~tissue) +
  scale_x_log10() +
  xlab("T>C conversion rate") +
  ylab("probability density") -> p

p %>% add_theme_SLAMtrack() 

```

# T>C conversion rates in known marker genes

The marker genes within `r exp_tissue_origin` are:  

```{r marker_genes_origin_table}

df_marker_origin %>% kable() %>% kable_styling() %>% scroll_box(width = "100%", height = "400px")

```

<br>
<br>

The marker genes within `r exp_tissue_target` are:  

```{r marker_genes_target_table}

df_marker_target %>% kable() %>% kable_styling() %>% scroll_box(width = "100%", height = "400px")

```

<br>
<br>

## Within the tissue of origin

Known marker genes within `r exp_tissue_origin`:  

```{r TC_marker_genes_origin}

df_tcounts %>% 
  filter(tissue == exp_tissue_origin) %>%
  merge_markers(df_marker_origin) %>% 
  plot_markers_points()

df_tcounts %>% 
  filter(tissue == exp_tissue_origin) %>%
  merge_markers(df_marker_origin) %>% 
  plot_markers_ridges()

```


## Within the target tissue

Known marker genes within `r exp_tissue_target`:  

```{r TC_marker_genes_target}

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  merge_markers(df_marker_target) %>% 
  plot_markers_points()

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  merge_markers(df_marker_target) %>% 
  plot_markers_ridges()

```


Known marker genes from `r exp_tissue_origin` within `r exp_tissue_target`:  

```{r TC_marker_genes_target_origin}

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  merge_markers(df_marker_origin) %>% 
  plot_markers_points()

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  merge_markers(df_marker_origin) %>% 
  plot_markers_ridges()

```