---
params:
 fname: "default"
 dir_data: "default"
 exp_type: "slamdunk" # slamdunk or smallslam

title: "SLAMtrack - Labelled genes"
subtitle: "`r params$fname`"
author: "RWH"
date: "`r format(Sys.time(), '%d %B %Y')`"

output: 
  html_document:
    theme: readable 
    highlight: pygments 
    anchor_sections: FALSE

---

```{r setup, include=FALSE}

if (params$exp_type == "slamdunk") {is_slamdunk <- TRUE} else {is_slamdunk <- FALSE}
dir_data <<- params$dir_data  ## <<- to make global variable
exp_type <<- params$exp_type

source(".Rprofile")
source("scripts/SLAMtrack_00_setup.R")
source("scripts/SLAMtrack_11_import_quick.R")

knitr::opts_chunk$set(
  echo = FALSE, 
  error = TRUE, 
  message = FALSE
  # fig.path = paste0(dir_data, "output/") - don't know why this not working
  )
  
```

# Approach 1 = threshold T>C conversion rate

## Identifying labelled reads in the tissue of origin

Using the negative control group, `r baseline_group`, we can define a threshold of T>C conversion that is likely to signify labelling.  This is set as any conversion rate exceeding a given centile (`r threshold_TC_centile`) in this control group.    

```{r set_TC_threshold_origin}

df_tcounts <- df_tcounts %>% drop_na(conversionRate)

df_tcounts %>% 
  filter(tissue == exp_tissue_origin) %>% 
  filter(group == baseline_group) %>% 
  select(conversionRate) -> baseline_conversionRates

baseline_conversionRates$conversionRate %>% 
  quantile(threshold_TC_centile) -> threshold_TC_absolute

df_tcounts %>% 
  filter(tissue == exp_tissue_origin) %>% 
  ggplot(aes(x = group, y = conversionRate)) +
  geom_violin(fill = NA) +  
  geom_jitter(alpha = 0.01) +
  scale_y_sqrt() +
  geom_hline(yintercept = threshold_TC_absolute, colour = "red", size = 0.5, linetype = 2) -> p

p %>% add_theme_SLAMtrack()

df_tcounts %>% 
  filter(tissue == exp_tissue_origin) %>% 
  mutate(
    conversionRate = conversionRate + cR_nudge
  ) %>% 
  ggplot(aes(y = 1, x = conversionRate, colour = group)) +
  geom_density_ridges(fill = NA) +
  scale_x_log10() +
  geom_vline(xintercept = threshold_TC_absolute, colour = "red", size = 0.5, linetype = 2) -> p

p %>% add_theme_SLAMtrack()

df_tcounts %>% 
  filter(tissue == exp_tissue_origin) %>% 
  mutate(
    conversionRate = conversionRate + cR_nudge
  ) %>% 
  ggplot(aes(y = library, x = conversionRate, fill = group)) +
  geom_density_ridges(alpha = 0.6) +
  scale_x_log10() +
  geom_vline(xintercept = threshold_TC_absolute, colour = "red", size = 0.5, linetype = 2) -> p

p %>% add_theme_SLAMtrack()

```

We can then classify each gene, for each library, as being either `labelled`, `not labelled`.  

```{r count_labelled_genes_origin}

df_tcounts %>% 
  filter(tissue == exp_tissue_origin) %>% 
  mutate(
    labelled = case_when(
      conversionRate < threshold_TC_absolute ~ "not labelled",
      conversionRate >= threshold_TC_absolute ~ "labelled"
    )) -> df_tcounts_origin

df_tcounts_origin %>% 
  group_by(sample_name) %>% 
  summarise(
    group = group[1],
    genes_all = length(labelled),
    genes_labelled = sum(labelled == "labelled"),
    pc = genes_labelled / genes_all
  ) %>% 
  arrange(group, sample_name) -> t

t %>% knitr::kable()

t %>% 
  group_by(group) %>% 
  summarise(
    percent_labelled = CI_95(pc, digit = 3)
  ) %>% knitr::kable()

t %>% 
  ggplot(aes(x = group, y = pc)) + 
  geom_boxplot(width = 0.2, alpha = 0.6, colour = "grey") +
  geom_jitter(width = 0.0, colour = "red") +
  ylab("percent genes labelled") -> p

p %>% add_theme_SLAMtrack()

```


How does labelling correlate with gene abundance?  Bin genes by abundance and plot % of labelled reads:

```{r plot_labelling_by_cpm_origin}

df_tcounts_origin %>% 
  mutate(reads_band = cut_width(readsCPM, width = 10)) %>% 
  group_by(group, library, reads_band) %>% 
  summarise(
    mean_CPM = mean(readsCPM),
    sum_CPM = sum(readsCPM),
    sum_lab = sum(labelled == "labelled"),
    total = length(labelled),
    pc = sum_lab / total
  ) %>% 
  mutate(
    cum_CPM = cumsum(sum_CPM)
    )-> df_tcounts_labelled_cut

df_tcounts_labelled_cut %>% 
  ggplot(aes(x = mean_CPM, y = pc, colour = group)) +
  geom_line(aes(group = library)) +
  scale_x_log10() +
  coord_cartesian(xlim = c(1, 1000)) +
  xlab("CPM") +
  ylab("Labelled reads (%)") -> p

p %>% add_theme_SLAMtrack()

df_tcounts_labelled_cut %>% 
  ggplot(aes(x = mean_CPM, y = pc, colour = group)) +
#  geom_line(aes(group = library)) +
  geom_smooth(fullrange = TRUE) +
  scale_x_log10() +
  coord_cartesian(xlim = c(2, 100)) +
  xlab("CPM") +
  ylab("Labelled reads (%)") -> p

p %>% add_theme_SLAMtrack()

```



## Identifying labelled reads in the target tissue

Using the negative control group, `r baseline_group`, we can define a threshold of T>C conversion that is likely to signify labelling.  This is set as any conversion rate exceeding a given centile (`r threshold_TC_centile`) in this control group.    

```{r set_TC_threshold_target}

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  filter(group == baseline_group) %>% 
  select(conversionRate) -> baseline_conversionRates

baseline_conversionRates$conversionRate %>% 
  quantile(threshold_TC_centile) -> threshold_TC_absolute

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  ggplot(aes(x = group, y = conversionRate)) +
  geom_violin(fill = NA) +  
  geom_jitter(alpha = 0.01) +
  scale_y_sqrt() +
  geom_hline(yintercept = threshold_TC_absolute, colour = "red", size = 0.5, linetype = 2) -> p

p %>% add_theme_SLAMtrack()

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  mutate(
    conversionRate = conversionRate + cR_nudge
  ) %>% 
  ggplot(aes(y = 1, x = conversionRate, colour = group)) +
  geom_density_ridges(fill = NA) +
  scale_x_log10() +
  geom_vline(xintercept = threshold_TC_absolute, colour = "red", size = 0.5, linetype = 2) -> p

p %>% add_theme_SLAMtrack()

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  mutate(
    conversionRate = conversionRate + cR_nudge
  ) %>% 
  ggplot(aes(y = library, x = conversionRate, fill = group)) +
  geom_density_ridges(alpha = 0.6) +
  scale_x_log10() +
  geom_vline(xintercept = threshold_TC_absolute, colour = "red", size = 0.5, linetype = 2) -> p

p %>% add_theme_SLAMtrack()

```

<br>
<br>

We can then classify each gene, for each library, as being either `labelled`, `not labelled`.  

```{r count_labelled_genes_target}

df_tcounts %>% 
  filter(tissue == exp_tissue_target) %>% 
  mutate(
    labelled = case_when(
      conversionRate < threshold_TC_absolute ~ "not labelled",
      conversionRate >= threshold_TC_absolute ~ "labelled"
    )) -> df_tcounts_target

df_tcounts_target %>% 
  group_by(sample_name) %>% 
  summarise(
    group = group[1],
    genes_all = length(labelled),
    genes_labelled = sum(labelled == "labelled"),
    pc = genes_labelled / genes_all
  ) %>% 
  arrange(group, sample_name) -> t

t %>% knitr::kable()

t %>% 
  group_by(group) %>% 
  summarise(
    percent_labelled = CI_95(pc, digit = 3)
  ) %>% knitr::kable()

t %>% 
  ggplot(aes(x = group, y = pc)) + 
  geom_boxplot(width = 0.2, alpha = 0.6, colour = "grey") +
  geom_jitter(width = 0.0, colour = "red") +
  ylab("percent genes labelled") -> p

p %>% add_theme_SLAMtrack()

```


How does labelling correlate with gene abundance?  Bin genes by abundance and plot % of labelled reads:

```{r plot_labelling_by_cpm_target}

df_tcounts_target %>% 
  mutate(reads_band = cut_width(readsCPM, width = 10)) %>% 
  group_by(group, library, reads_band) %>% 
  summarise(
    mean_CPM = mean(readsCPM),
    sum_CPM = sum(readsCPM),
    sum_lab = sum(labelled == "labelled"),
    total = length(labelled),
    pc = sum_lab / total
  ) %>% 
  mutate(
    cum_CPM = cumsum(sum_CPM)
    )-> df_tcounts_labelled_cut

df_tcounts_labelled_cut %>% 
  ggplot(aes(x = mean_CPM, y = pc, colour = group)) +
  geom_line(aes(group = library)) +
  scale_x_log10() +
  coord_cartesian(xlim = c(1, 1000)) +
  xlab("CPM") +
  ylab("Labelled reads (%)") -> p

p %>% add_theme_SLAMtrack()

df_tcounts_labelled_cut %>% 
  ggplot(aes(x = mean_CPM, y = pc, colour = group)) +
#  geom_line(aes(group = library)) +
  geom_smooth(fullrange = TRUE) +
  scale_x_log10() +
  coord_cartesian(xlim = c(2, 100)) +
  xlab("CPM") +
  ylab("Labelled reads (%)") -> p

p %>% add_theme_SLAMtrack()

```


We can then write lists of the genes with `r exp_tissue_target` that are:

- present and ALWAYS labelled within `r exp_tissue_origin` 
- present and SOMETIMES labelled within `r exp_tissue_origin` 
- present and NEVER labelled within `r exp_tissue_origin` 
- NOT present within `r exp_tissue_origin`

...and then compare labelling rates within these groups.  


```{r classify_genes_in_target}

df_tcounts_origin %>% 
  filter(readsCPM > 0) %>% 
  group_by(gene_name, group) %>% 
  summarise(
    n_labelled = sum(labelled == "labelled"),
    n_unlabelled = sum(labelled == "not labelled")
  ) %>% 
  mutate(
    n = n_labelled + n_unlabelled,
    classification = case_when(
      n == n_labelled ~ "always",
      n == n_unlabelled ~ "never"
    )
  ) %>% 
  select(gene_name, group, classification) %>% 
  pivot_wider(names_from = group, values_from = classification) -> df_labelling_in_origin
  
df_labelling_in_origin %>% 
  mutate(
    labelling_in_origin = case_when(
      !!as.name(pos_group) == "always" & !!as.name(neg_group) == "never" ~ "robust positive",
      !!as.name(pos_group) == "never" ~ "robust negative",
      !!as.name(neg_group) != "never" ~ "leaky"
    )) %>% 
  select(gene_name, labelling_in_origin) -> df_labelling_in_origin_summary

df_labelling_in_origin_summary$labelling_in_origin %>% 
  NA_sub_x(paste0("not classified in ", exp_tissue_origin)) ->
  df_labelling_in_origin_summary$labelling_in_origin

df_tcounts_target %>% 
  left_join(df_labelling_in_origin_summary, by = c("gene_name")) -> df_tcounts_target

df_tcounts_target$labelling_in_origin %>% 
  NA_sub_x(paste0("not present in ", exp_tissue_origin)) %>% 
  fct_relevel(
    "robust positive", 
    "robust negative", 
    "leaky"
    ) -> df_tcounts_target$labelling_in_origin


```


The number of genes in each category is:  

```{r show_genes_in_target}

new_name = paste0("labelling in ", exp_tissue_origin)

colnames(df_tcounts_target)[which(colnames(df_tcounts_target) == "labelling_in_origin")] <- new_name

df_tcounts_target %>% 
  group_by(sample_name, !!as.name(new_name)) %>% 
  summarise(n = length(gene_name)) %>% 
    group_by(!!as.name(new_name)) %>% 
    summarise(
      median = median(n),
      Q1 = quantile(n, 0.25),
      Q3 = quantile(n, 0.75)
    ) %>% knitr::kable(caption = paste0("number of genes in each category within each ", exp_tissue_target, " library"))

```


Does rate of T>C conversion differ between these groups?

```{r plot_genes_in_target_by_category}

df_tcounts_target %>% 
  mutate(
    conversionRate = conversionRate + cR_nudge
  ) %>% 
    ggplot(aes(y = 1, x = conversionRate, colour = group)) +
    geom_density_ridges(fill = NA) +
    scale_x_log10() +
    facet_wrap(as.formula(paste0("~ ", "`", new_name, "`"))) +
    xlab("T>C conversion rate") +
    ylab("") +
    ggtitle(paste0(
      "conversion rates in ", 
      exp_tissue_target, 
      "\n(stratified by labelling status of gene in ",
      exp_tissue_origin,
      ")"
    )) -> p

p <- p %>% add_theme_SLAMtrack()
p + theme_smaller()

df_tcounts_target %>% 
  filter(group %in% c(pos_group, neg_group)) %>% 
  mutate(
    conversionRate = conversionRate + cR_nudge
  ) %>% 
    ggplot(aes(y = 1, x = conversionRate, colour = group)) +
    geom_density_ridges(fill = NA) +
    scale_x_log10() +
    facet_wrap(as.formula(paste0("~ ", "`", new_name, "`")))  +
    xlab("T>C conversion rate") +
    ylab("") +
    ggtitle(paste0(
      "conversion rates in ", 
      exp_tissue_target, 
      "\n(stratified by labelling status of gene in ",
      exp_tissue_origin,
      ")"
    )) -> q

q <- q %>% add_theme_SLAMtrack()
q + theme_smaller()

```



# Approach 2 = gene-wise delta T>C conversion rate

## In the tissue of origin

```{r delta_origin}

df_tcounts %>%
  filter(tissue == exp_tissue_origin) %>% 
  filter(group == baseline_group) %>% 
  group_by(gene_name) %>% 
  summarise(
    mean_CPM = mean(readsCPM)
    ) -> df_tcounts_origin_cpm

df_tcounts %>%
  filter(tissue == exp_tissue_origin) %>% 
  group_by(gene_name, group) %>% 
  summarise(
    mean_CR = mean(conversionRate)
    ) %>% 
  pivot_wider(names_from = group, values_from = mean_CR) %>% 
  mutate(delta := !!as.name(pos_group) - !!as.name(neg_group)) %>% 
  left_join(df_tcounts_origin_cpm) %>% 
  mutate(
    `delta sign` = case_when(
      delta > 0 ~ "pos", 
      delta < 0 ~ "neg", 
      delta == 0 ~ "zero")
  ) -> df_tcounts_delta_origin

df_tcounts_delta_origin %>% 
  filter(mean_CPM > 5) %>% 
  ggplot(aes(x = mean_CPM, y = delta)) + 
  geom_point(alpha = 0.05) +
  geom_hline(yintercept = 0, colour = "Red", linetype = 2) +
  scale_x_log10() +
  theme_SLAMtrack()

df_tcounts_delta_origin %>% 
  drop_na() %>% 
  group_by(`delta sign`) %>% 
  summarise(count = n()) %>% 
  knitr::kable()

df_tcounts_delta_origin %>% 
  filter(mean_CPM > 1) %>% 
  filter(abs(delta) < 0.1) %>% 
  ggplot(aes(x = delta, fill = `delta sign`)) +
  geom_histogram() + 
  facet_wrap(~`delta sign`) -> p

p %>% add_theme_SLAMtrack() + theme_smaller()

df_tcounts_delta_origin %>% 
  merge_markers_delta(df_marker_origin) %>% 
  plot_markers_points_delta()

```

## In the target tissue

```{r delta_target}

df_tcounts %>%
  filter(tissue == exp_tissue_target) %>% 
  filter(group == baseline_group) %>% 
  group_by(gene_name) %>% 
  summarise(
    mean_CPM = mean(readsCPM)
    ) -> df_tcounts_target_cpm

df_tcounts %>%
  filter(tissue == exp_tissue_target) %>% 
  group_by(gene_name, group) %>% 
  summarise(
    mean_CR = mean(conversionRate)
    ) %>% 
  pivot_wider(names_from = group, values_from = mean_CR) %>% 
  mutate(delta := !!as.name(pos_group) - !!as.name(neg_group)) %>% 
  left_join(df_tcounts_target_cpm) %>% 
  mutate(
    `delta sign` = case_when(
      delta > 0 ~ "pos", 
      delta < 0 ~ "neg", 
      delta == 0 ~ "zero")
  ) -> df_tcounts_delta_target

df_tcounts_delta_target %>% 
  filter(mean_CPM > 5) %>% 
  ggplot(aes(x = mean_CPM, y = delta)) + 
  geom_point(alpha = 0.05) +
  geom_smooth() +
  geom_hline(yintercept = 0, colour = "Red", linetype = 2) +
  scale_x_log10() +
  theme_SLAMtrack()

df_tcounts_delta_target %>% 
  drop_na() %>% 
  group_by(`delta sign`) %>% 
  summarise(count = n()) %>% 
  knitr::kable()

df_tcounts_delta_target %>% 
  filter(mean_CPM > 1) %>% 
  filter(abs(delta) < 0.005) %>% 
  ggplot(aes(x = delta, fill = `delta sign`)) +
  geom_histogram() + 
  facet_wrap(~`delta sign`) -> p

p %>% add_theme_SLAMtrack() + theme_smaller()

df_tcounts_delta_target %>% 
  merge_markers_delta(df_marker_origin) %>% 
  plot_markers_points_delta(cpm_threshold = 10, delta_threshold = 0.05)

df_tcounts_delta_target %>% 
  merge_markers_delta(df_marker_target) %>% 
  plot_markers_points_delta(delta_threshold = 0.05)

  
```

