---
params:
 fname: "default"
 dir_data: "default"
 exp_type: "slamdunk" # slamdunk or smallslam

title: "SLAMtrack - Labelled genes"
subtitle: "`r params$fname`"
author: "RWH"
date: "`r format(Sys.time(), '%d %B %Y')`"

output: 
  html_document:
    theme: readable 
    highlight: pygments 
    anchor_sections: FALSE

---

```{r setup, include=FALSE}

if (params$exp_type == "slamdunk") {is_slamdunk <- TRUE} else {is_slamdunk <- FALSE}
dir_data <<- params$dir_data  ## <<- to make global variable
exp_type <<- params$exp_type

source(".Rprofile")
source("scripts/SLAMtrack_00_setup.R")
source("scripts/SLAMtrack_11_import_quick.R")
source("scripts/SLAMtrack_21_labelled_genes.R")

knitr::opts_chunk$set(
  echo = FALSE, 
  error = TRUE, 
  message = FALSE
  # fig.path = paste0(dir_data, "output/") - don't know why this not working
  )
  
```

# Finding labelled genes

In order to identify labelled genes, calculate gene-wise delta T>C conversion rates.  In other words, for each gene find the difference between the median T>C conversion rate in a positive group (e.g. Cre pos) and a control group (e.g. Cre neg).  

```{r get_deltas}

df_tcounts %>% 
  get_delta(exp_tissue = exp_tissue_origin, 
            group1 = baseline_group, 
            group2 = neg_group) -> df_delta_origin_baseline

df_tcounts %>% 
  get_delta(exp_tissue = exp_tissue_origin, 
            group1 = neg_group, 
            group2 = pos_group) -> df_delta_origin

df_tcounts %>% 
  get_delta(exp_tissue = exp_tissue_target, 
            group1 = neg_group, 
            group2 = pos_group) -> df_delta_target

```

<br>
<br>

## Basal rates

First compare the `r baseline_group` and `r neg_group` groups to establish distribution of deltas under basal conditions.  Set a threshold at the `r threshold_delta_basal_centile` centile, above which a read is considered labelled in `r exp_tissue_origin`.  

```{r baseline_rates}

Qx_origin_baseline <- df_delta_origin_baseline$delta %>% quantile(threshold_delta_basal_centile)

df_delta_origin_baseline %>% delta_tabulate()
df_delta_origin_baseline %>% delta_plot_distrubution(Qx_origin_baseline) -> p
df_delta_origin_baseline %>% delta_plot_vs_cpm() -> q

p$plot
q$plot_full

```

<br>
<br>

## Labelling in tissue of origin  

Then compare the `r pos_group` and `r neg_group` in order to determine which genes are labelled in `r exp_tissue_origin` under experimental conditions.  

<br>

Classify genes in `r exp_tissue_origin` as being one of:    

- definitely labelled (delta exceeds `r threshold_delta_basal_centile` centile in basal conditions)  
- definitely NOT labelled (delta less then `r 1 - threshold_delta_basal_centile` centile in basal conditions)  
- possibly labelled (delta between these thresholds)  
- not abundant (median CPM below `r threshold_median_cpm_abundant`) - this trumps other three criteria  


```{r call_labelled_genes}

df_delta_origin %>% 
  mutate(
    labelled_in_origin = case_when(
      median_CPM >= threshold_median_cpm_abundant & delta >= Qx_origin_baseline ~ paste0("definitely labelled in ", exp_tissue_origin),
      median_CPM >= threshold_median_cpm_abundant & delta <= -Qx_origin_baseline ~ paste0("definitely not labelled in ", exp_tissue_origin),
      median_CPM >= threshold_median_cpm_abundant & delta > -Qx_origin_baseline & delta < Qx_origin_baseline ~ paste0("possibly labelled in ", exp_tissue_origin),
      median_CPM < threshold_median_cpm_abundant ~ paste0("not abundant in ", exp_tissue_origin)
    )
  ) %>% 
  select(gene_name, labelled_in_origin) -> df_genes_labelled_in_origin

df_delta_origin %>% 
  left_join(df_genes_labelled_in_origin) -> df_delta_origin

df_delta_target %>% 
  left_join(df_genes_labelled_in_origin) -> df_delta_target

df_delta_target$labelled_in_origin %>% 
  NA_sub_x(paste0("not present in ", exp_tissue_origin)) -> df_delta_target$labelled_in_origin

```

<br>
<br>

NB The distribution of gene abundance in `r exp_tissue_origin` is:

```{r median_cpm_threshold}

df_delta_origin %>% 
  ggplot(aes(x = median_CPM)) + 
  geom_histogram(alpha = 0.6, fill = "red") + 
  geom_vline(aes(xintercept = threshold_median_cpm_abundant), linetype = 2, colour = "blue") +
  scale_x_log10() +
  theme_SLAMtrack()  

```

<br>
<br>

Number of genes so classified in `r exp_tissue_origin`:  

```{r show_labelled_genes_origin}

df_genes_labelled_in_origin %>% 
  group_by(labelled_in_origin) %>% 
  summarise(n = length(gene_name)) %>% 
  knitr::kable()

```


Number of genes so classified - i.e. according to labelling status in `r exp_tissue_origin` in `r exp_tissue_target`:

```{r show_labelled_genes_target}
df_delta_target %>% 
  group_by(labelled_in_origin) %>% 
  summarise(n = length(gene_name)) %>% 
  knitr::kable()

```

<br>
<br>

## Delta TC in tissue of origin

Plotting gene-wise delta TC in `r exp_tissue_origin`:  

```{r show_delta_origin}

df_delta_origin %>% delta_tabulate()

df_delta_origin %>% delta_plot_distrubution(Qx_origin_baseline) -> p
p$plot
p$plot + facet_wrap(~labelled_in_origin)

df_delta_origin %>% delta_plot_rug() 

df_delta_origin %>% delta_plot_vs_cpm() -> q
q$plot_full

```

<br>
<br>

### Delta TC in known marker genes

Within `r exp_tissue_origin`, what is gene-wise delta TC and labelling status within known marker genes?  

```{r delta_by_marker_gene}

df_delta_origin %>% 
  delta_merge_markers(df_marker_origin) %>% 
  delta_plot_markers_points()

df_delta_origin %>% 
  delta_merge_markers(df_marker_origin) %>% 
  delta_plot_markers_ridges()

df_delta_origin %>% 
  delta_merge_markers(df_marker_origin) %>% 
  group_by(cell, labelled_in_origin) %>% 
  summarise(n = length(gene_name)) %>% 
  pivot_wider(names_from = labelled_in_origin, values_from = n) %>% 
  NA_sub_zero() %>% 
  knitr::kable()

```

<br>
<br>

## Delta TC in target tissue

Look within those reads not present at all in `r exp_tissue_origin`; set a threshold at centile `r threshold_delta_target_centile`, above which gene called as being labelled in `r exp_tissue_target`.  

```{r call_labelled_genes_in_target}

df_delta_target %>% 
  filter(labelled_in_origin == paste0("not present in ", exp_tissue_origin)) %>% 
  select(delta) -> df_delta_target_not_in_origin

Qx_target_not_in_origin <- df_delta_target_not_in_origin$delta %>% quantile(threshold_delta_target_centile)

```

<br>
<br>

Then plot gene-wise delta TC in `r exp_tissue_target`:  

```{r show_delta_target}

df_delta_target %>% delta_tabulate()

df_delta_target %>% delta_plot_distrubution(Qx_target_not_in_origin) -> p
p$plot
p$plot + facet_wrap(~labelled_in_origin)

df_delta_target %>% delta_plot_rug() 

df_delta_target %>% delta_plot_vs_cpm() -> q
q$plot_full

```

<br>
<br>

## Final list of transferred genes

Using these criteria, we can write a list of genes that are likely to have been transferred from `r exp_tissue_origin` to `r exp_tissue_target`.   

```{r list_labelled_genes}

df_delta_target %>% 
  filter(labelled_in_origin == paste0("definitely labelled in ", exp_tissue_origin)) %>% 
  filter(delta > Qx_target_not_in_origin) %>% 
  select(gene_name, delta) -> df_transferred_genes

```


There are `r nrow(df_transferred_genes)` genes in this list:  

```{r show_labelled_genes}

df_transferred_genes %>% knitr::kable()

```

